Parkinson's Disease cluster analysis

### 2016-06-29 writeup/manuscript/manuscript.tex
Update...a lot of things

Various formal measures were used to determine the optimal number of clusters for the dataset. The
optimal $k$ according to the Gap Statistic and the 1-standard-error method suggested by
Tibshirani\cite{tibshirani01gap} was $k = 4$ (Supplementary Figure~\ref{fig:gap}). Other cluster validation
methods (within sum of squares scree plot, minimum average silhouette width) suggested $k =
2, 3, 4$, where $k = 2, 3$ simply divided the data uninformatively into groups with varying levels
of overall PD severity. Thus $k = 4$ was selected to offer a good blend of model fit,
informativeness, and parsimony.

@@

\begin{table*}[t]
  \centering
  \caption{Domains clustering summary. Unless otherwise specified, statistics are
  reported as mean (sd).}
  \label{tab:nmd}
  \begin{threeparttable}
    \small
    \begin{tabular}{lrllll}
    \toprule
    & Cluster & 1 & 2 & 3 & 4 \\
    & $n$ & 406 (45\%) & 189 (21\%) & 221 (24\%) & 88 (10\%) \\
    \midrule
    Nonmotor domains & Cardiovascular & 0.7 (1.6)\tnote{24} & 2.3 (2.9)\tnote{134} & 1.1 (2.0)\tnote{24} &
    7.0 (6.0)\tnote{123} \\
    & Sleep/fatigue & 4.3 (4.8)\tnote{234} & 14.9 (8.2)\tnote{134} & 6.9 (6.3)\tnote{124}
    & 21.0 (10.0)\tnote{123} \\
  & Mood/apathy & 3.1 (4.5)\tnote{234} & 16.1 (14.0)\tnote{134} & 6.6
    (8.1)\tnote{124} & 23.4 (14.3)\tnote{123} \\
    & Perception/hallucination & 0.5 (1.7)\tnote{24} & 1.9 (3.5)\tnote{134} & 0.8
    (2.0)\tnote{24} & 8.6 (6.9)\tnote{123} \\
    & Attention/memory & 2.8 (4.2)\tnote{24} & 8.9 (8.1)\tnote{134} & 3.2 (4.4)\tnote{24} &
    15.6 (11.1)\tnote{123} \\
    & Gastrointestinal & 2.8 (3.9)\tnote{234} & 8.8 (7.1)\tnote{134} & 4.1
    (4.7)\tnote{124} & 14.6 (9.6)\tnote{123} \\
    & Urinary & 4.5 (5.9)\tnote{234} & 12.4 (9.7)\tnote{134} & 6.1 (6.4)\tnote{124} &
    20.3 (9.9)\tnote{123} \\
    & Sexual function & 1.5 (3.1)\tnote{24} & 6.4 (7.4)\tnote{134} & 2.5 (4.1)\tnote{24} &
    9.0 (9.7)\tnote{123} \\
    & Miscellaneous & 3.9 (4.7)\tnote{24} & 13.2 (8.6)\tnote{13} & 5.2 (5.8)\tnote{24} &
    14.0 (9.5)\tnote{13} \\
    \midrule
    Motor symptoms & Tremor & 1.9 (1.8)\tnote{34} & 1.6 (1.9)\tnote{34} & 4.3 (2.8)\tnote{124} & 3.5
    (3.8)\tnote{123} \\
    & Bradykinesia & 1.5 (0.9)\tnote{234} & 2.1 (1.1)\tnote{134} & 3.6 (0.9)\tnote{124} &
    4.0 (1.5)\tnote{123} \\
    & Rigidity & 1.5 (0.9)\tnote{234} & 1.8 (1.1)\tnote{134} & 3.3 (0.9)\tnote{124} &
    3.8 (1.4)\tnote{123} \\
    & Axial & 1.8 (1.6)\tnote{234} & 3.3 (2.1)\tnote{134} & 4.3 (2.4)\tnote{124} &
    7.4 (2.9)\tnote{123} \\
    \midrule
    Variables not & Sex (\% male) & 63 & 53\tnote{3} & 71\tnote{2} & 58 \\
    used in analysis & CISI total & 5.7 (3.2)\tnote{234} & 9.2 (3.9)\tnote{14} & 9.7
    (3.5)\tnote{14} & 14.8 (5.1)\tnote{123} \\
    & Age (34--89) & 62.7 (9.7)\tnote{34} & 64 (9.0)\tnote{4} & 65 (10.2)\tnote{14} & 70.5 (8.9)\tnote{123} \\
    & PD onset & 56 (10.6) & 55.2 (10.1) & 56.9 (11.1) & 58.2 (11.4) \\
    & PD duration & 6.7 (4.8)\tnote{234} & 8.7 (5.7)\tnote{14} & 8 (5.5)\tnote{14} &
    12.3
    (8.1)\tnote{123} \\
    \bottomrule
  \end{tabular}
  \begin{tablenotes}
    \small
    \item[1] Significant difference with cluster 1 (corrected $p < 0.05$)
    \item[2] Significant difference with cluster 2 (corrected $p < 0.05$)
    \item[3] Significant difference with cluster 3 (corrected $p < 0.05$)
    \item[4] Significant difference with cluster 4 (corrected $p < 0.05$)
  \end{tablenotes}
  \end{threeparttable}
\end{table*}

@@

Most variables had little to no correlation with PD duration (Supplementary Figure~\ref{fig:corr}). Variables
with the highest correlation include symptoms urinary frequency, swallowing, and drooling; domains
miscellaneous, urinary, and gastrointestinal; and total CISI score. Scatterplots for CISI Total,
Tremor, Anxiety, and Depression appear in Supplementary Figure~\ref{fig:long4}.

>>>>

Various formal measures were used to determine the optimal number of clusters for the dataset. The
optimal $k$ according to the Gap Statistic and the 1-standard-error method suggested by
Tibshirani\cite{tibshirani01gap} was $k = 4$ (Supplementary Figure~\ref{fig:gap}). Other cluster validation
methods (within sum of squares error scree plot, minimum average silhouette width) suggested $k =
2, 3, 4$, where $k = 2, 3$ simply divided the data uninformatively into groups with varying levels
of overall PD severity. Thus $k = 4$ was selected to offer a good blend of model fit,
informativeness, and parsimony.

@@

\begin{table*}[t]
  \centering
  \caption{Domains clustering summary. Unless otherwise specified, statistics are
  reported as mean (sd).}
  \label{tab:nmd}
  \begin{threeparttable}
    \small
    \begin{tabular}{lrllll}
    \toprule
    	 Cluster 	 1 	 2 	 3 	 4 \\
    	 $n$ 	 406 (45\%) 	 189 (21\%) 	 221 (24\%) 	 88 (10\%) \\
    \midrule
    Nonmotor domains 	 Cardiovascular 	 0.7 (1.6)\tnote{24} 	 2.3 (2.9)\tnote{134} 	 1.1 (2.0)\tnote{24} 	
    7.0 (6.0)\tnote{123} \\
    a	 Sleep/fatigue 	 4.3 (4.8)\tnote{234} 	 14.9 (8.2)\tnote{134} 	 6.9 (6.3)\tnote{124} 21.0 (10.0)\tnote{123} \\
  	 a  Mood/apathy 	 3.1 (4.5)\tnote{234} 	 16.1 (14.0)\tnote{134} 	 6.6 (8.1)\tnote{124} 	 23.4 (14.3)\tnote{123} \\
    	 Perception/hallucination 	 0.5 (1.7)\tnote{24} 	 1.9 (3.5)\tnote{134} 	 0.8 (2.0)\tnote{24} 	 8.6 (6.9)\tnote{123} \\
    	 Attention/memory 	 2.8 (4.2)\tnote{24} 	 8.9 (8.1)\tnote{134} 	 3.2 (4.4)\tnote{24} 	15.6 (11.1)\tnote{123} \\
    	 Gastrointestinal 	 2.8 (3.9)\tnote{234} 	 8.8 (7.1)\tnote{134} 	 4.1 (4.7)\tnote{124} 	 14.6 (9.6)\tnote{123} \\
    	 Urinary 	 4.5 (5.9)\tnote{234} 	 12.4 (9.7)\tnote{134} 	 6.1 (6.4)\tnote{124} 	20.3 (9.9)\tnote{123} \\
    	 Sexual function 	 1.5 (3.1)\tnote{24} 	 6.4 (7.4)\tnote{134} 	 2.5 (4.1)\tnote{24} 	9.0 (9.7)\tnote{123} \\
    	 Miscellaneous 	 3.9 (4.7)\tnote{24} 	 13.2 (8.6)\tnote{13} 	 5.2 (5.8)\tnote{24} 	14.0 (9.5)\tnote{13} \\
    \midrule
    Motor symptoms 	 Tremor 	 1.9 (1.8)\tnote{34} 	 1.6 (1.9)\tnote{34} 	 4.3 (2.8)\tnote{124} 	 3.5 (3.8)\tnote{123} \\
    	 Bradykinesia 	 1.5 (0.9)\tnote{234} 	 2.1 (1.1)\tnote{134} 	 3.6 (0.9)\tnote{124} 	4.0 (1.5)\tnote{123} \\
    	 Rigidity 	 1.5 (0.9)\tnote{234} 	 1.8 (1.1)\tnote{134} 	 3.3 (0.9)\tnote{124} 	3.8 (1.4)\tnote{123} \\
    	 Axial 	 1.8 (1.6)\tnote{234} 	 3.3 (2.1)\tnote{134} 	 4.3 (2.4)\tnote{124} 	7.4 (2.9)\tnote{123} \\
    \midrule
    Variables not 	 Sex (\% male) 	 63 	 53\tnote{3} 	 71\tnote{2} 	 58 \\
    used in analysis 	 CISI total 	 5.7 (3.2)\tnote{234} 	 9.2 (3.9)\tnote{14} 	 9.7 (3.5)\tnote{14} 	 14.8 (5.1)\tnote{123} \\
    	 Age (34--89) 	 62.7 (9.7)\tnote{34} 	 64 (9.0)\tnote{4} 	 65 (10.2)\tnote{14} 	 70.5 (8.9)\tnote{123} \\
    	 PD onset 	 56 (10.6) 	 55.2 (10.1) 	 56.9 (11.1) 	 58.2 (11.4) \\
    	 PD duration 	 6.7 (4.8)\tnote{234} 	 8.7 (5.7)\tnote{14} 	 8 (5.5)\tnote{14} 	12.3 (8.1)\tnote{123} \\
    \bottomrule
  \end{tabular}
  \begin{tablenotes}
    \small
    \item[1] Significant difference with cluster 1 (corrected $p < 0.05$)
    \item[2] Significant difference with cluster 2 (corrected $p < 0.05$)
    \item[3] Significant difference with cluster 3 (corrected $p < 0.05$)
    \item[4] Significant difference with cluster 4 (corrected $p < 0.05$)
  \end{tablenotes}
  \end{threeparttable}
\end{table*}

@@

Most variables had little to no correlation with PD duration (Supplementary Figure~\ref{fig:corr}).
Variables with the highest correlation included symptoms urinary frequency, swallowing, and
drooling; domains miscellaneous, urinary, and gastrointestinal; and total CISI score. Scatterplots
for CISI Total, Tremor, Anxiety, and Depression appear in Supplementary Figure~\ref{fig:long4}.
Total CISI score increased predictably for all four clusters with increasing PD duration, but
tremor, anxiety, and depression had no significant longitudinal progression; indeed, for cluster
2, mood/apathy symptoms appeared to be highest at PD onset, decreasing as time progressed. Notably,
for the motor/nonmotor-dominant groups, differences in severity for motor and nonmotor symptoms
respectively was present early at PD onset, persisting throughout the range of PD durations
observed.
